---
title: "twoxtwoxkTables Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{twoxtwoxkTables Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(twoxtwoxkTables)
```

# Introduction

Commonly used in epidemiology, 2x2xk tables are a way to express the relationship between a binary independent variable (often an exposure or treatment) and a binary dependent variable (often an outcome or disease).
This package provides functions that generate common epidemiological measures for 2x2xk tables.  

In this vignette, I demonstrate the usage of the functions in this package and provide a comparison to existing functions published on CRAN.

This package is not published to CRAN, but can be installed by running the line `devtools::install_github("smeah/twoxtwoxkTables")` in the R console.  To install with the vignette, run ``devtools::install_github("smeah/twoxtwoxkTables", build_vignettes = TRUE)`.

This package was created by Sabir Meah (smeah@umich.edu) as a part of BIOSTAT-625: Computing with Big Data at the University of Michigan.

Functions:

* cond.odds.ratios() - Conditional Odds Ratios

Calculates the conditional odds ratios for each of the k strata in a 2x2xk table.

* common.odds.ratio() - Common Odds Ratios

Calculates the common odds ratio across all k strata in a 2x2xk table.

* mantelhaenzel.test() - Cochran–Mantel–Haenszel Test

Performs the Cochran–Mantel–Haenszel chi-squared test (two-sided without continuity correction) on a 2x2xk table.  
Tests the null hypothesis that the common odds ratio is 1 against the alternative hypothesis that it is not one.

# Usage





